---
title: "Homework 7"
author: "Mari Sanders"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(nlme)
library(lme4)
data <- read_csv("HW7-politeness_data.csv") 
```

# Problem 1 

## (a)

```{r}
data %>% ggplot(aes(x = gender, y = frequency)) + 
  geom_boxplot() + facet_wrap(~attitude)

data %>% 
  group_by(gender, attitude) %>% 
  summarize(median = median(frequency))

```

In general, males have a lower frequency than females. 

## (b)

### Model 
```{r}
LMM1 <- lme(frequency ~ gender + attitude, random = ~1 | subject,  data = data, method='REML')

summary(LMM1)

```

### Covariance Matrix Structure Individual Participant

```{r}
sigma_u2 <- as.numeric(VarCorr(LMM1)[1, "Variance"])    
sigma_e2 <- as.numeric(VarCorr(LMM1)[2, "Variance"])   


subject_id <- "F1"

n_i <- sum(data$subject == subject_id)

I <- diag(n_i)
J <- matrix(1, n_i, n_i)

Cov_Yi <- sigma_u2 * J + sigma_e2 * I
Cov_Yi
```

Since the covariance of the non diagonal terms are all the same, the covariance structure for an individual participant is compound symmetry. 

### Covariance Structure Fixed Effects 

```{r}
vcov(LMM1)
```

### BLUPs

```{r}
random.effects(LMM1)
```

### Residuals 

```{r}
resid_cond <- resid(LMM1, type = "response")
resid_marg <- resid(LMM1, type = "normalized")  
data$resid_cond <- resid_cond
data$resid_marg <- resid_marg
head(data[, c("subject", "frequency", "resid_cond", "resid_marg")])
```

```{r}
ggplot(data, aes(x = subject, y = resid_cond)) +
  geom_boxplot(fill = "skyblue") +
  labs(title = "Conditional Residuals by Participant",
       x = "Participant (subject)", y = "Conditional Residuals") +
  theme_minimal()

ggplot(data, aes(x = subject, y = resid_marg)) +
  geom_boxplot(fill = "lightpink") +
  labs(title = "Marginal Residuals by Participant",
       x = "Participant (subject)", y = "Marginal Residuals") +
  theme_minimal()
```

## (c)

```{r}
LMM2 <- lme(frequency ~ gender + attitude + gender*attitude, random = ~1 | subject,  data = data, method='REML')

summary(LMM2)

LMM1.comp <- lme(frequency ~ gender + attitude, random = ~1 | subject,  data = data, method='ML') 
LMM2.comp <- lme(frequency ~ gender + attitude + gender*attitude, random = ~1 | subject,  data = data, method='ML')
anova(LMM2.comp,LMM1.comp) 
```

Since the p-value is 0.2392, we fail to reject the null hypothesis and say that the simpler model is better. 

## (d)

```{r}
LMM3 <- lmer(frequency ~ gender + attitude + (1 | subject) + (1 | scenario), data = data)

summary(LMM3)

```

## Covariance Matrix for an Individual Participant

```{r}
id <- "F1"
df_i <- data[data$subject == id, ]
n_i <- nrow(df_i)
vc <- as.data.frame(VarCorr(LMM3))
sigma_subject <- vc$vcov[vc$grp == "subject"]
sigma_scenario <- vc$vcov[vc$grp == "scenario"]
sigma_resid <- sigma(LMM3)^2
I <- diag(n_i)
J <- matrix(1, n_i, n_i)
scenario_vec <- df_i$scenario
B <- outer(scenario_vec, scenario_vec, FUN = Vectorize(function(a, b) as.numeric(a == b)))
Cov_Yi <- sigma_subject * J + sigma_scenario * B + sigma_resid * I
Cov_Yi
```

Formal speech is associated with pitch that is 20 Hz lower on average than informal speech, when adjusting for gender and random effects. 